---
title: "Analysis of White Wines"
author: "Kevin Turner"
date: "January 24, 2019"
output: html_document
---

***

```{r include=FALSE, global_options}
# set global options for Knitr
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# load packages
library(ggplot2)
library(knitr)
library(gridExtra)
library(tidyr)
library(dplyr)
library(ggcorrplot)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, load_the_data}
# load data
whiteWines <- read.csv("wineQualityWhites.csv")

# remove column 'X' as it is not needed
whiteWines$X <- NULL
```

The data set contains 4,898 white wines with 11 attributes on quantifying
the chemical properties of each wine and 1 attribute providing a quality score 
as determined by at least 3 expert wine tasters. Quality was rated on a scale of
0 to 10.

Let's have a look at the attributes available within this dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, display_attributes}
#display the attributes in the data set
names(whiteWines)
```

#### Description of attributes:

   1. fixed acidity (tartaric acid - g / dm^3): most acids involved with wine
   are fixed or nonvolatile (do not evaporate readily)

   2. volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in
   wine, which at too high of levels can lead to an unpleasant, vinegar taste

   3. citric acid (g / dm^3): found in small quantities, citric acid can add
   'freshness' and flavor to wines

   4. residual sugar (g / dm^3): the amount of sugar remaining after 
   fermentation stops, it's rare to find wines with less than 1 gram/liter and
   wines with greater than 45 grams/liter are considered sweet

   5. chlorides (sodium chloride - g / dm^3): the amount of salt in the wine

   6. free sulfur dioxide (mg / dm^3): the free form of SO2 exists in 
   equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion;
   it prevents microbial growth and the oxidation of wine

   7. total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02;
   in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

   8. density (g / cm^3): the density of wine is close to that of water
   depending on the percent alcohol and sugar content

   9. pH: describes how acidic or basic a wine is on a scale from 0 (very 
   acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

   10. sulphates (potassium sulphate - g / dm3): a wine additive which can
   contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial
   and antioxidant

   11. alcohol (% by volume): the percent alcohol content of the wine

   12. quality (score between 0 and 10) - Output variable (based on sensory 
   data)

# Univariate Plots

Now that we know more about the attributes involved with our dataset, let's have 
a look at how these attributes are structered.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_variables}
# display strucure of data
str(whiteWines)
```

Most of the variables are of the numeric type, which makes sense as the data
represents qauntative measurements. The last variable, Quality, is of the type
integer since values were recorded as whole numbers on a scale of 0 to 10.

Next, let's take a statistical look at the values.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_summary}
# display statistical summary of data
summary(whiteWines)
```

After reviewing the statistical information there are a few interesting 
observations. Some of the attributes have large gaps between thier 3rd
Quartile and Max values. We'll look a bit more at those in a minute.

First, lets take a quick view of how the wines rated by quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_quality}
# create bar chart of quality values
ggplot(whiteWines, aes(quality)) + geom_bar(fill='darkgreen')
```

The quality of the wines we are anlyzing look to follow a nice bell curve with
most wines falling in the middle of the scale and a one or two on each end.

Now lets see a quick visual overview of the rest of our data points.



```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_all}
# create key, value pairs to compre quality
ww_long <- gather(whiteWines, 
                  key = measure,value = value, -'quality')

# create facet containing histograms for each attribute
ggplot(ww_long, aes(value)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_histogram(fill = 'darkblue')
```


From the histograms above we can see that some of our data is close to normal
distribution while other data is right skewed. There are no attributes that are
left skewed in the dataset. We can aso deduce that there are outliers in several
attributes based on the bin widths and what appears to be large empty space to
the right of the bars.

Lets have a closer look at a __chlorides__ and __free.sulfur.dioxide__, two of
the attributes that appear to have extreme outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_chlorides}
# create histogram of chlorides
ggplot(whiteWines, aes(chlorides)) + 
  geom_histogram(fill = "darkslategrey") +
  
  # calculate outliers and label them on histogram
  stat_bin(geom="text", aes(label = ifelse( 
    x > 3 * IQR(whiteWines$chlorides) + quantile(whiteWines$chlorides)[4] & 
      ..count.. > 0, ..count.., "")), vjust = -1.5)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_free_sulfur}
# create histogram of free sulfur dioxide
ggplot(whiteWines, aes(free.sulfur.dioxide)) + 
  geom_histogram(fill = "darkslategrey") +
  
  # calculate outliers and label them on histogram
  stat_bin(geom="text", aes(
    label = ifelse(x > 3 * IQR(whiteWines$free.sulfur.dioxide) + 
    quantile(whiteWines$free.sulfur.dioxide)[4] & 
      ..count.. > 0, ..count.., "")), vjust = -1.5)
```

In both the above histograms the major outliers have been labeled to make them 
easier to identify. Major outliers are calculated based on 
Third Quartile + 3 * Inner Quartile Range.

The below histogram is for __pH__, an attribute that more closely follows a 
normal distribution. As evident by the lack of any labeled bins, there were no 
major outliers found here.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_pH}
# create histogram for pH
ggplot(whiteWines, aes(pH)) + 
  geom_histogram(fill = "darkslategrey") +
  
  # calculate outliers and label them on histogram
  stat_bin(geom="text", aes(
    label = ifelse(x > 3 * IQR(whiteWines$pH) + quantile(whiteWines$pH)[4] & 
      ..count.. > 0, ..count.., "")), vjust = -1.5)
```


# Univariate Analysis

The dataset contains 4,898 observations and 12 variables. 11 variables are of
the numeric data type and 1 variable is an integer.

The main feature of interest in this dataset would be the Quality score. 
Throughout this analysis we will investigate how the other variables 
contribute, positive or negative, to that quality score.

In order to organize the data in a way that was easier to display as an overview
the tidyr library was used to create a new dataframe. Utilizing this library 
value:key pairs were created for each variable and value. This process is
defined as taking "wide"" data, many columns and fewer rows, and making it 
"long", many rows and fewer columns. The resulting dataframe's structure 
contains 53,878 observations and 3 variables.

Most distributions with in this dataset can be described as right skewed, or
close to a normal. Several attributes in the dataset do contain major outliers,
however the values are within an acceptable range for the process so for the
purpose of this analysis they will be included.

# Bivariate Plots

Let's see if we can make any determinations as to whether any of these
attributes have an affect on the quality of the wines.

We'll start by looking at a visual overview of how each attribute value
correlates to the quality rating. We'll use scatterplots and smooth curves to
help us more easuly identify trends.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate_plot_quality_all}
# create facet containing histograms for each attribute
ggplot(ww_long, aes(value, quality)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_point(color = 'darkblue') +
    geom_smooth(color = 'orange')
```

Very interesting. It appears that some attributes  like __alcohol__,
__sulphates__, and __pH__ may have very little affect on the quality of the wine
as evident by the relativly flat smooth curve.

However, a few atributes do stand out. __free.sulfur.dioxide__ and 
__total.sulfur.dioxide__ appear to have a sweet spot for higher quality scores
and values outside that range, lower or higher, seems to negatively affect
quality. Additionally, it appears the the higher the __volatile.acidity__ the
lower the quality of wine. This makes sense since too high of levels of 
__volatile.acidity__ can lead to an unpleasant, vinegar taste.

We are going to take a closer look at these attributes.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate_plot_free.sulfur}
ggplot(whiteWines, aes(free.sulfur.dioxide, quality)) +
  geom_jitter(color = 'blue') +
  geom_smooth(color = 'orange', fill = 'yellow')
```

Here we can more clearly see that there is a negative trend with
__free.sulfur.dioxide__ and __quality__. When __free.sulfur.dioxide__ values are
with in a specific range of about 0 - 100, quality does not seem affected much,
however once the value start getting beyond 100 there is a noticable degredation
in the __quality__ of the wine.

So what has an effect on __free.sulfur.dioxide__? Let's again take a visual
overview of our attributes, but this time focusing on how they compare to
__free.sulfur.dioxide__.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate_plot_free.sulfur_all}
# create key, value pairs to compare free sulfur dioxide
ww_free.sulfur.dioxide <- gather(whiteWines, 
                  key = measure,value = value, 
                  -'quality', -'free.sulfur.dioxide')

# create facet containing histograms for each attribute
ggplot(ww_free.sulfur.dioxide, aes(free.sulfur.dioxide, value)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_point(color = 'darkblue') +
    geom_smooth(color = 'orange')
```

Clearly more __free.sulfur.dioxide__ leads to more __total.sulfur.dioxide__.
This makes perfect sense. There are a couple other trends that look interesting,
__pH__ and __sulphates__, but at first glance it seems that the outliers may 
have some some influence here.

Since there is a strong relationship between __free.sulfur.dioxide__ and 
__total.sulfur.dioxide__, lets explore it a little more.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate_plot_total.sulfur}
# create scatter plot with smooth curve for free sulfur dioxide
ggplot(whiteWines, aes(total.sulfur.dioxide, free.sulfur.dioxide)) +
  geom_jitter(color = 'blue') +
  geom_smooth(color = 'orange', fill = 'yellow')
```

Looking at this relationship more closely we can see that even excluding the
outlier we have a postive relationsip between __free.sulfur.dioxide__ and
__total.sulfur.dioxide__.

But does this mean anything? Let's keep going by looking at how 
__total.sulfur.dioxide__ is affected by the other attributes.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate_plot_total.sulfur_all}
# create key, value pairs to compare total sulfur dioxide
ww_total.sulfur.dioxide <- gather(whiteWines, 
                  key = measure,value = value, 
                  -'quality', -'total.sulfur.dioxide')

# create facet containing histograms for each attribute
ggplot(ww_total.sulfur.dioxide, aes(total.sulfur.dioxide, value)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_point(color = 'darkblue') +
    geom_smooth(color = 'orange')
```

of course we can see the relationship between __free.sulfur.dioxide__ and
__total.sulfur.dioxide__ as we explored earlier. Also seen earlier with
__free.sulfur.dioxide__ we can see that there is a positive trend with 
__sulphates__ a depicted below.


```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate_plot_sulphates}
# create scatter plot with smooth curve for total sulfur dioxide
ggplot(whiteWines, aes(total.sulfur.dioxide, sulphates)) +
  geom_jitter(color = 'blue') +
  geom_smooth(color = 'orange', fill = 'yellow')
```

Since __sulphates__ is a wine additive which can contribute to sulfur dioxide 
gas (S02) levels it seems logical that the more __sulphates__ found in the wine
then higher levels of __free.sulfur.dioxide__ and __total.sulfer.dioxide__ will
exist.

# Bivariate Analysis

We explored the relationship between seraval attributes through this bivariate
analysis.

Beginning with an overview of how each attribute affected __qaulity__, we
identified a couple attributes to take a deeper look at. From there we explored
__free.sulfur.dioxide__, which led is to __total.sulfur.dioxide__, then finally
to __sulphates__. 

We explored how each attribute affected the next and can
conclude that the more __sulphates__ added to a wine increases the 
__free.sulfur.dioxide__ and __total.sulfur.dioxie__ levels which can have a
negative impact on the quality of wine.

# Multivariate Plots Section

In previous sections of the analysis we explored our data set and the indvidual
attributes it contains. We also looked at how attrubites affect other 
attributes. Now let see multiple attributes together. The first thing we will 
look at is how each attribute correlates with the others.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
# create correlation for all attributes
corr <- round(cor(whiteWines), 1)

ggcorrplot(corr, p.mat = cor_pmat(whiteWines),
           hc.order = TRUE, type = "lower",
           color = c("red", "white", "blue"),
           outline.col = "white", lab = TRUE)


```

So our dataset contains some postive corellations, identified in purple, and
some negative correlations, identified in red. We can see the strongest postive
correlation is between __residial.sugar__ and __denstiy__. The strongest
negative correlation is between __alcohol__ and __density__. There are also a 
number of non-significant correlations marke by the X over the box.

# Multivariate Analysis

It's interesting to see that there are many more positive correlations between
attributes within the dataset than negative ones. As certain levels of some
attributes rise it directtly affects these other attributes in a positive way.
As an example, though this probably seems intuitive, if there are more 
__residual.sugar__ then it is expected that the wine will have more __density__.

As the __Quality Score__ was the primary focus of this dataset, it was a little
surprising to see that __citric.acid__, something I previously thought was a
large component in the taste of wine, has no real correlation to the 
__quality__.

------

# Final Plots and Summary

We've spent a bit of time exploring the dataset. We learned a lot and came to
the conclusion that __Quality Score__ was the main attribute to focus on.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# create facet containing histograms for each attribute
ggplot(ww_long, aes(value)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_histogram(fill = 'darkblue')
```

We began by taking a look at the details of the other attributes through an 
overview of thier histograms.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# create facet containing histograms for each attribute
ggplot(ww_long, aes(value, quality)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_point(color = 'darkblue') +
    geom_smooth(color = 'orange')
```

We  continued our analysis by looking at how these attributes affected our 
__Quality Score__ by looking at scatterplots and smooth curves.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# create correlation for all attributes
ggcorrplot(corr, p.mat = cor_pmat(whiteWines),
           hc.order = TRUE, type = "lower",
           color = c("red", "white", "blue"),
           outline.col = "white", lab = TRUE)

```

Finally we looked at the correlation between each attribute and the other
attributes in the dataset.

------

# Reflection

Working with this dataset and the R language was a fun experience. Understanding
what graphs and charts would represent the data appropriatly was a challenge at 
times.

It may be possible to create a predicitive model of creating __quality__ wine
from the data avaialbe. It would certainly be more benificial if the dataset
contained much more data to analyze.
