---
title: "Analysis of White Wines"
author: "Kevin Turner"
date: "January 24, 2019"
output: html_document
---

***

```{r include=FALSE, global_options}
# set global options for Knitr
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# load packages
library(ggplot2)
library(knitr)
library(gridExtra)
library(tidyr)
library(dplyr)
library(ggcorrplot)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, load_the_data}
# load data
whiteWines <- read.csv("wineQualityWhites.csv")

# remove column 'X' as it is not needed
whiteWines$X <- NULL
```

The data set contains 4,898 white wines with 11 attributes on quantifying
the chemical properties of each wine and 1 attribute providing a quality score 
as determined by at least 3 expert wine tasters. Quality was rated on a scale of
0 to 10.

Let's have a look at the attributes available within this dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, display_attributes}
#display the attributes in the data set
names(whiteWines)
```

#### Description of attributes:

   1. fixed acidity (tartaric acid - g / dm^3): most acids involved with wine
   are fixed or nonvolatile (do not evaporate readily)

   2. volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in
   wine, which at too high of levels can lead to an unpleasant, vinegar taste

   3. citric acid (g / dm^3): found in small quantities, citric acid can add
   'freshness' and flavor to wines

   4. residual sugar (g / dm^3): the amount of sugar remaining after 
   fermentation stops, it's rare to find wines with less than 1 gram/liter and
   wines with greater than 45 grams/liter are considered sweet

   5. chlorides (sodium chloride - g / dm^3): the amount of salt in the wine

   6. free sulfur dioxide (mg / dm^3): the free form of SO2 exists in 
   equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion;
   it prevents microbial growth and the oxidation of wine

   7. total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02;
   in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

   8. density (g / cm^3): the density of wine is close to that of water
   depending on the percent alcohol and sugar content

   9. pH: describes how acidic or basic a wine is on a scale from 0 (very 
   acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

   10. sulphates (potassium sulphate - g / dm3): a wine additive which can
   contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial
   and antioxidant

   11. alcohol (% by volume): the percent alcohol content of the wine

   12. quality (score between 0 and 10) - Output variable (based on sensory 
   data)

# Univariate Plots

Now that we know more about the attributes involved with our dataset, let's have 
a look at how these attributes are structered.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_variables}
# display strucure of data
str(whiteWines)
```

Most of the variables are of the numeric type, which makes sense as the data
represents qauntative measurements. The last variable, Quality, is of the type
integer since values were recorded as whole numbers on a scale of 0 to 10.

Next, let's take a statistical look at the values.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_summary}
# display statistical summary of data
summary(whiteWines)
```

After reviewing the statistical information there are a few interesting 
observations. Some of the attributes have large gaps between thier 3rd
Quartile and Max values. We'll look a bit more at those in a minute.

First, lets take a quick view of how the wines rated by quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_quality}
# create bar chart of quality values
ggplot(whiteWines, aes(quality)) + geom_bar(fill='darkgreen')
```

The quality of the wines we are anlyzing look to follow a nice bell curve with
most wines falling in the middle of the scale and a one or two on each end.

Now lets see a quick visual overview of the rest of our data points.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_all}
# create key, value pairs to compre quality
ww_long <- gather(whiteWines, 
                  key = measure,value = value, -'quality')

# create facet containing histograms for each attribute
ggplot(ww_long, aes(value)) + 
    facet_wrap(~measure, scales = 'free') + 
    geom_histogram(fill = 'darkblue')
```

From the histograms above we can see that some of our data is close to normal
distribution while other data is right skewed. There are no attributes that are
left skewed in the dataset. We can also deduce that there are outliers in several
attributes based on the bin widths and what appears to be large empty space to
the right of the bars.

Lets have a closer look at a __chlorides__ and __free.sulfur.dioxide__, two of
the attributes that appear to have extreme outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_chlorides}
# create boxplot and histogram sidy by side for comparison
grid.arrange(ggplot(whiteWines, aes(1, chlorides)) +
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = 'red'),
             ggplot(whiteWines, aes(chlorides)) +
               geom_histogram(), ncol = 2) 
             
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_free_sulfur}
# create boxplot and histogram sidy by side for comparison
grid.arrange(ggplot(whiteWines, aes(1, free.sulfur.dioxide)) +
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = 'red'),
             ggplot(whiteWines, aes(free.sulfur.dioxide)) +
               geom_histogram(),ncol = 2)
```

So our dateset contains quite a few outliers. These outliers could affect the
outcome or quality of our analysis. So we will create a new dataframe that
excludes all the rows of data that contains outiers in any of the attributes.


```{r, echo=FALSE, message=FALSE, warning=FALSE, remove_outliers}
# create new dataframe to remove outliers
ww_no_outliers <- whiteWines

# use boxplot to identify outliers then remove those rows from dataframe
outlier_removal <- function(x){
  outliers <- boxplot(ww_no_outliers[,x], plot = FALSE)$out
  if (length(outliers)){
    ww_no_outliers <- ww_no_outliers[-which(ww_no_outliers[,x] %in% outliers),]
    return(ww_no_outliers)
  }
  else {
    return(ww_no_outliers)
  }
}

# loop through dataframe to remove outliers from columns 1 through 11
for (name in names(ww_no_outliers[1:11])) {
  for (column in ww_no_outliers[name]) {
    if (is.numeric(column)) {
      ww_no_outliers <- outlier_removal(name)
    }
  }
}


```

This new dataframe contains 3,973 observations. 925 rows that contained
outliers were removed.

```{r, echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_variables2}
# display structure of dataframe
str(ww_no_outliers)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_summary2}
# display summary of dataframe
summary(ww_no_outliers)
```

Lets have another look at __chlorides__ and __free.sulfur.dioxide__ after the
outliers have been removed.

```{r, echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_chlorides2}
# create boxplot and histogram sidy by side for comparison
grid.arrange(ggplot(ww_no_outliers, aes(1, chlorides)) +
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = 'red'),
             ggplot(ww_no_outliers, aes(chlorides)) +
               geom_histogram(), ncol = 2)
             
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, univariate_plot_free_sulfur2}
# create boxplot and histogram sidy by side for comparison
grid.arrange(ggplot(ww_no_outliers, aes(1, free.sulfur.dioxide)) +
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = 'red'),
             ggplot(ww_no_outliers, aes(free.sulfur.dioxide)) +
               geom_histogram(), ncol = 2)
```

We can see that the data is much cleaner and easier to visualize now. For the
rest of this anlysis we will be using the data that excludes the outliers.

# Univariate Analysis

The orignal dataset contained 4,898 observations and 12 variables. 
11 variables are of the numeric data type and 1 variable is an integer. A subset
of this data was created to exclude outliers for the analysis. The subsetted
data contains the same 12 variables, but only includes 3,815 observations.

The main feature of interest in this dataset would be the Quality score. 
Throughout this analysis we will investigate how the other variables 
contribute, positive or negative, to that quality score.

In order to organize the data in a way that was easier to display as an 
overview, the tidyr library was used to create a new dataframe. Utilizing this
library, value:key pairs were created for each variable and value. This process
is defined as taking "wide"" data, many columns and fewer rows, and making it 
"long", many rows and fewer columns. The resulting dataframe's structure 
contains 53,878 observations and 3 variables.

Most distributions with in this dataset can be described as right skewed, or
close to a normal.

# Bivariate Plots

Let's see if we can make any determinations as to whether any of these
attributes have an affect on the __quality__ of the wines.

We'll start by looking for any correlations between each of the attributes.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate1}
# create correlation for all attributes
corr <- round(cor(ww_no_outliers), 1)

ggcorrplot(corr, p.mat = cor_pmat(ww_no_outliers),
           hc.order = TRUE, type = "lower",
           color = c("red", "white", "blue"),
           outline.col = "white", lab = TRUE)


```

A correlation coefficient of > 4.0 is consindered a relatively strong 
relationship. Between 2.0 and 4.0 is considered moderate and anything less than
2.0 is weak.

This is pretty interesting. It appears that the strongest correlations are
between __density__ and both __alcohol__ and __residual.sugar__. There also 
looks to be some moderate correlations between our main attribute of interest, 
__quality__, and the attributes __alcohol__, __density__, __chlorides__, and 
__total.sulfur.dioxide__.

Lets explore these relationships further.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate2}
# function to create scatter plots with trend lines
create_scatter <- function(x, y, xlabel, ylabel){
  ggplot(ww_no_outliers, aes(x, y)) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE, size=1) +
  xlab(xlabel) +
  ylab(ylabel)
}

# create scatterplot by function
create_scatter(ww_no_outliers$density, ww_no_outliers$alcohol, 'density',
               'alcohol')
```

The relationship between __alcohol__ and __denisty__ is a negative relationship.
This suggest that the higher the __alcohol__ content of the wine, the less 
__dense__ the wine will be.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate3}
# create scatterplot by function
create_scatter(ww_no_outliers$density, ww_no_outliers$residual.sugar, 'density',
               'residual.sugar')
```

Conversely, the relationship between __residual.sugar__ and __density__ is a
postive. This makes sense since it would seem logical that the more __sugar__
that exists in the wine the more __dense__ it would become.

So what does the relationship between __residual.sugar__ and __alcohol__ look
like?
```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate4}
# create scatterplot by function
create_scatter(ww_no_outliers$residual.sugar, ww_no_outliers$alcohol, 
               'residual.sugar', 'alcohol')
```

The negative relationship we see above coincides with the previous relationships
we explored. A lower __residual.sugar__ value tends to result in a higher 
__alcohol__ content.

How about __quality__? Let's look at some of the moderate relationships between
__quality__ and the other attributes.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate5}
# function to create boxplot with jitter
bivariate_quality <- function(y, ylabel){
  ggplot(ww_no_outliers, aes(factor(quality), y)) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  stat_summary(fun.y = 'mean',
               geom = 'point',
               color = 'red',
               shape = 8,
               size = 4) +
  ylab(ylabel)
}

# create graph by function
bivariate_quality(ww_no_outliers$alcohol, 'alcohol')
```

We've already determined that no strong relationships exist between the
rated __quality__ of the wine and any other attributes. At best there are a few
moderate relationships. When comparing __quality__ to __alcohol__, we see that
amongst the wines rated between 5 and 8 there exists __alcohol__ values from the
whole range in the dataset. In fact the highest rated wines contain only higher
alcohol percentages. This observation may just be a result of a relatively small
dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate6}
# create scatterplot by function
bivariate_quality(ww_no_outliers$density, 'density')
```

As with __alcohol__, __denstiy__ values are spread accross each of the
__quality__ levels. The highest quality level only contains lower __density__.
This coincides with the relationship between __desnity__ and __alcohol__ that
we looked at earlier.

Lets have have a look at the last two attributes with a moderate relationship, 
__chlorides__ and __total.sulfur.dioxide__. I suspect we will see the same trend
though.

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate7}
# create scatterplot by function
bivariate_quality(ww_no_outliers$chlorides, 'chlorides')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, bivariate8}
# create scatterplot by function
bivariate_quality(ww_no_outliers$total.sulfur.dioxide, 'total.sulfur.dioxide')
```

There is a definite trend here. Though moderate relationships exist between
__quality__ and other attributes, the relationships do not appear to be 
definitive except for the highest rated wines. Again, this may just be a result
of relatively small dataset.

# Bivariate Analysis

We explored the relationship between seraval attributes through this bivariate
analysis.

Begining with the srongest relationship between attributes 
 __alcohol__/ __density__ and __residual.sugar__/ __density__ , 
 then looking at our __quality__ attribute and it's moderate relationships.


Through this analysis I had hoped we would identify correlations between
attributes and it's related __quality__ score, however given the data it's not 
possible to derive how each attribute could be manipulated to affect the 
quality of wine.

# Multivariate Plots Section

Let's take a look at the scatter plots we created earlier and overlay 
__quality__ values on them.

```{r echo=FALSE, message=FALSE, warning=FALSE, multivariate1}
# function to create scatter with quality overlay
multivariate_scatter <- function(x, y, xlabel, ylabel){
  ggplot(ww_no_outliers, aes(x, y, color = factor(quality))) +
    geom_point(alpha = 0.8, size = 1) +
    geom_smooth(method = "lm", se = FALSE,size=1)  +
    scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
    xlab(xlabel) +
    ylab(ylabel)
}

# create scatterplot by function
multivariate_scatter(ww_no_outliers$density, ww_no_outliers$alcohol, 'density',
               'alcohol')
```

As before, we see the negative relationship between __alcohol__ and __density__,
but now it is much clearer that each __quality__ value covers the whole range
of of values in the other attributes, but the highest __quality__ of 9 only
spans over a small set of values in the other attributes.

```{r echo=FALSE, message=FALSE, warning=FALSE, multivariate2}
# create scatterplot by function
multivariate_scatter(ww_no_outliers$density, ww_no_outliers$residual.sugar, 
                     'density', 'residual.sugar')
```

An interesting observation in this graph and the previous one is that the
__quality__ value of 9 has an oposite trend as the others. That's to say that if
the the trend is postive for __quality__ values 3 through 8, then 9 is negative.

```{r echo=FALSE, message=FALSE, warning=FALSE, multivariate3}
# create scatterplot by function
multivariate_scatter(ww_no_outliers$residual.sugar, ww_no_outliers$alcohol, 
               'residual.sugar', 'alcohol')
```

Again we can see the same trend where only the highest __quality__ value of 9 is
limited to a small range in the other attributes values and the __quality__
value of 9 follows and oppisite trend than the rest  of the __quality__ values.

# Multivariate Analysis

The most intersting finding from the multivariate analysis would have relate to 
the __quality__ attribute. When __quality__ is analyzed along with other
attributes each level of quality follows the same trend , postive or negative,
with the eception of the highest __quality__ rating of 9. The highest
__quality__ rating of 9 consistently trends opposite of the other ratings.

------

# Final Plots and Summary

We've spent a bit of time exploring the dataset. We learned a lot and came to
the conclusion that even though __Quality Score__ appeared to be a good starting
focus as a main attribute, there was not a lot of evidence that any one or 
combination of atributes directly contributed to it.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# create facet containing histograms for each attribute
grid.arrange(ggplot(whiteWines, aes(1, chlorides)) +
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = 'red') +
               ggtitle('Data With Outliers') +
               theme(plot.title = element_text(hjust = 0.5),
                     axis.title.x=element_blank(),
                     axis.text.x=element_blank(),
                     axis.ticks.x=element_blank()) +
               ylab('Chlorides (sodium chloride - g / dm^3)'),
             ggplot(ww_no_outliers, aes(1, chlorides)) +
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = 'red')  +
               ggtitle('Data Without Outliers') +
               theme(plot.title = element_text(hjust = 0.5),
                     axis.title.x=element_blank(),
                     axis.text.x=element_blank(),
                     axis.ticks.x=element_blank()) +
               ylab('Chlorides (sodium chloride - g / dm^3)'), ncol = 2) 
```

When we began our analysis we found that our dataset contained a lot of
outliers. In order to mitigate the effect that had on our analysis, a new subset
of the data was created to remove those outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# create correlation for all attributes
corr <- round(cor(ww_no_outliers), 1)

ggcorrplot(corr, p.mat = cor_pmat(ww_no_outliers),
           hc.order = TRUE, type = "lower",
           color = c("red", "white", "blue"),
           outline.col = "white", lab = TRUE) +
   ggtitle('Correlations Between Attributes') +
   theme(plot.title = element_text(hjust = 0.5))
  
```

After ensuring all outliers were removed from the dataset, we had a look at the 
correlation between each of the attributes to determine is there were any
relationships beyond __quality__ that were of intrest to look at.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
multivariate_scatter(ww_no_outliers$residual.sugar, ww_no_outliers$alcohol, 
               'residual.sugar', 'alcohol') +
  ggtitle('Quality of Wine as it relates to % of Alcohol and Amount of 
          Residual Sugars') +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab('Residual Sugar (g / dm^3)') +
  ylab('Alcohol (% by volume)')

```

Finally we explored some of the attribute through bivarite and multivariate
analysis.

------

# Reflection

Working with this dataset and the R language was a fun experience. Understanding
what graphs and charts would represent the data appropriatly was a challenge at 
times.

It may be possible to create a predicitive model of creating __quality__ wine
from the data avaialbe. It would certainly be more benificial if the dataset
contained much more data to analyze.
